Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-12-18 manu-draft.qmd
Abstract and Intro up to Line 74 for manu-draft.qmd
Initial edits up to line 74.

Climate change is an existential threat to environmental resources that sustain human health and the environment. In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth. Changing physical conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress. Three datasets of varying sample design and coverage were evaluated to characterize physical environments in Tampa Bay and the potential relationships with seagrass change. Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 ppt per year. These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay segments during the summer months. Simple linear models provided limited evidence that the recent seagrass decline is associated with hotter and fresher conditions. Trends in warming and increased precipitation in the region will continue, further creating suboptimal conditions for seagrasses in Tampa Bay. These results provide a motivating example for resource managers to consider the likelihood of reduced resilience of estuarine resources due to shifting ecological baselines from climate change. Conventional management approaches that focus on limiting nutrient loads should be continued, but the effectiveness may be limited and additional actions will be needed to maintain ecosystem health in the future.

@@

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable, if not more, than other critical coastal environments [@duarte2008a], which represents an existential threat to the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007], whereas natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create supportive conditions for growth. A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014]. There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources. Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014] and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in seagrass areal coverage through sustained and long-term reductions in nutrients. Cooperation among management, regulatory, public, and private sectors were critical aspects of both examples [@tango2016; @sherwood2016]. In other cases, the reversal of seagrass losses through nutrient reductions alone may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016]. Costly alternative management actions, in addition to nutrient reductions, are needed in these situations. Management actions are much more effective in estuaries that are stressed but not past critical thresholds that define an alternative state.

>>>>

Climate change is an existential threat to environmental resources that sustain human health and the environment. In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth. Changing physical water quality conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress. Three datasets of varying sample design and coverage were evaluated to characterize physical environments in Tampa Bay and the potential relationships with seagrass change. Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 PSU per year. These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay segments during the summer months. Simple linear models provided limited evidence that recent seagrass declines are associated with hotter and fresher conditions. Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay. These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines from climate change could complicate long-standing management paradigms. While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be limited and warrant additional, complementary interventions to maintain ecosystem health into the future.

@@

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable, if not more significant, than other critical coastal environments [@duarte2008a], which represents an existential threat to the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007]. Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Contemporary management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

The sustained coverage or restoration of seagrasses in stressed coastal environments requires management actions that support growth and coverage expansion. A long-standing approach adopted by numerous management entities has been the control of external nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014]. There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources. Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014] and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in seagrass areal coverage through sustained and long-term reductions in external nutrient loads. Cooperation among management, regulatory, public, and private sectors were critical aspects of both examples [@tango2016; @sherwood2016]. In other cases, the reversal of seagrass losses through nutrient reductions alone may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016]. Complementary management actions, in addition to nutrient reductions, are needed in these situations. Management actions with co=benefits are much more effective in estuaries that are stressed but not past critical thresholds that produce an alternative ecosystem state.
